PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-05-10 papers/NAR_web_server_2009/paper.sup.tex
Update of manuscript for final submission

% Using the following to mark changes for review
\usepackage{color}
\definecolor{red}{rgb}{1,0,0}
\newcommand{\mc}[1]{\textcolor{red}{#1}}
%\newcommand{\mc}[1]{#1}   % Replace above line with this to eliminate all coloring

@@

@@

\begin{table}[tbp]
\begin{center}
\begin{tiny}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline   & PEACE & 0.871 & 0.809 & 0.186 & 0.049 & 129 & 4 & 129 \\
\cline{2-9}  MetaSim 454 & WCD(1) & 0.073 & 0.073 & 0.014 & 0.082 & 7028.500 & 5296.533 & 74 \\
\cline{2-9}  Simulation & WCD(2) &  0.020 & 0.020 & 0.008 & 0.198 & 13085 & 10610 & 64 \\
\cline{2-9}  & Cap3 & 0.005 & 0.005 & 0.005 & 0.565 & 19222 & 15425 & 132 \\
\hline
\hline Metasim  & PEACE & 0.583 & 0.372 & 0.230 & 0.111 & 392.914 & 149.257 & 3454 \\
\cline{2-9} Illumina  & Cap3 & 0.006 & 0.006 & 0.002 & 1.766 & 51663.057 & 30107.429 & 291 \\
\hline
\end{tabular}
\end{tiny}
\end{center}
\caption{Comparisons of runs on on short read data, as generated by
  \metasim \cite{Richter2008}; each figure is an average over 30 runs.
  For \wcd(1) we use wcd default parameters, while for \wcd(2) we set
  it to match the \peace\/ parameters ($w=75$, $T=60$); \wcd\/ was
  unable to return any results for Illumina data, hence has been
  omitted.  Each 454 simulation produced 25,000 reads from the 100
  zebra-fish transcripts, ranging  in size from 200bp to 270bp (aveage = 250bp,
  standard deviation = 17bp).  Each Illumina simulation produce 150K
  reads from the 100 zebra-fish transcript, each exactly 62bp.}\label{SRSim}
\end{table}

>>>>

% Using the following to mark changes for review
\usepackage{color}
\definecolor{red}{rgb}{1,0,0}
%\newcommand{\mc}[1]{\textcolor{red}{#1}}
\newcommand{\mc}[1]{#1}   % Replace above line with this to eliminate all coloring

@@

\mc{All runs were conducted on 3.0 GHz Intel Xeon EM64T CPUs with 2 MB
caches and a 800 MHz front side bus, model number Xeon LV 3.0 (2005)}

@@

\begin{table}[htbp]
\begin{center}
\begin{tiny}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline   454 & PEACE & 0.871 & 0.809 & 0.186 & 0.049 & 129 & 4 & 129 \\
\cline{2-9}    Simulation & WCD(1) & 0.459 & 0.448 & 0.946 & 0.003 & 599 & 235 & 76 \\
\cline{2-9}  & TGICL & 0.297 & 0.288 & 0.984 & 0.005 & 886 & 0 & 68 \\
\hline
\hline Illumina         & PEACE & 0.384 & 0.343 & 0.230 & 0.018 & 673  & 299 & 3463 \\
\cline{2-9} Simulation  & TGICL & 0.131 & 0.127 & 0.977 & 0.008 & 1522 & 0   & 373 \\
\hline
\end{tabular}
\end{tiny}
\end{center}
\caption{Comparisons of runs on on short read data, as generated by
  \metasim \cite{Richter2008}; each figure is an average over 30 runs.
  For \wcd we use parameters suggested by the authors  (``-H 1 -T 160''). \wcd\/ was
  unable to return any reasonable results for Illumina data, while
  \capthree\/ was unable to return results for either sumulation.  Each
  454 simulation produced 25,000 reads from the 100 zebra-fish
  transcripts, ranging  in size from 200bp to 270bp (aveage = 250bp, 
  standard deviation = 17bp).  Each Illumina simulation produce 150K
  reads from the 100 zebra-fish transcript, each exactly 62bp.}\label{SRSim}
\end{table}

### 2010-05-10 papers/NAR_web_server_2009/paper.tex
Update of manuscript for final submission

Given a set of transcript fragments sampled from across the genome, a
necessary first step of the set's analysis is that of clustering: separating
the fragments according to the transcript from which they were
derived.  Frequently performed implicitly by assembly tools,
clustering the data as a ``pre-assembly'' step has a number of
advantages.  Most significantly: performing this step will
allow the application of the assembly tool to individual clusters --
saving significant amounts of time \cite{Hazelhurst08a}.

However, clustering is a computationally challenging problem.  Even
with the smaller number of ESTs produced using Sanger Sequencing, the
runtime and memory requirements to cluster on the basis of pair-wise
sequence alignments make such an approach infeasible in practice.  The
much larger data-set size produced by NGS technologies exacerbates
this problem.  To deal with this, \peace\/ combines our own version of
the $d^2$ alignment-free sequence distance function \cite{Hide94} and
the concept of a {\it minimum spanning tree} \cite{Prim57} to quickly
and accurately find clusters of ESTs expressed from the same gene
without reference to a sequenced genome.  \mc{Compared against
  clustering tool in the literature
  \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
    Pertea2003,Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be of
  equivalent quality to the \wcd\/ tool \cite{Hazelhurst08a}, more
  sensitive and robust than the other tools, and more user friendly
  than any published clustering tool.  Nor are any of} \mc{these tools
  designed for the ease of installation and use that \peace\/
  provides.}

@@

\mc{In applying the tools to real data Sanger data, we used the Human
  Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
  the A076941 {\it Arabidopsis thaliana} data set used to test wcd\/
  \cite{Hazelhurst08a,Hazelhurst08b} (see Table~\ref{SangerReal}).
  We notice essentially identical results for \peace\/ and wcd\/ in
  quality, both significantly better than \capthree\/ in Sensitivity and Type
  1 error rate, while slightly worse in the Jaccard Index and Type 2
  error.  In runtime we see some inconsistency, with \peace\/ showing a
  60\% runtime improvement over \wcd\/ in the first data set, but requiring 20\%
  more time than \wcd\/ in the Arabidopsis data set.  (We cannot make
  a reasonable runtime comparison to \capthree, as the runtime of that
  tool reflects both clustering and assembly.)  In Table~S1 of the
  Supplementary Materials we present runtimes of several more sets,
  observing that while \peace\/ appears to be significantly 
  faster on the smaller sets, \wcd\/ does overtake it for larger
  sets.}

@@

\mc{We tested the three tools on short-read data using the \metasim\/
  tool of {\it Richter et al.} \cite{Richter2008}.  Encoded into
  \metasim\/ are sequence generation and error models corresponding to
  several technologies, including the 454 short-read technology
  (producing reads ranging in size from 200bp to 370bp with a mean of
  250bp and a standard deviation of 17bp), and the Illumina short-read
  technology (producing reads of exactly 62bp).  Experiments were run
  using two sets of \wcd\/ parameters (the default parameters, and
  setting window size to 75 and threshold to 75 -- thus matching the
  equivalent \peace\/ parameters), and using default values
  of \capthree.  Only with \peace\/ were we able to get reasonable
  results, seeing a sensitivity of 0.871 for the 454 reads and
  0.583 or the shorter Illumina reads, a Jaccard Index of 0.809 and
  0.372 respectively, and other comparably figures.  (See Table~S3 in
  the Supplementary Materials.)  In short, \peace performs quite well
  for 454 reads and provides some useful information for Illumina
  reads, while \wcd\/ and \capthree\/ are unable to produce working
  results.}

>>>>

Given a set of transcript fragments sampled from across the genome, a
necessary first step of the set's analysis is clustering: separating
the fragments according to the transcript from which they were
derived.  Frequently performed implicitly by assembly tools,
clustering the data as a ``pre-assembly'' step has a number of
advantages.  Most significantly: performing this step will
allow the application of the assembly tool to individual clusters --
saving significant amounts of time \cite{Hazelhurst08a}.

Clustering is a computationally challenging problem; the runtime and
memory requirements to cluster on the basis of pair-wise sequence
alignments make such an approach infeasible in practice.  To deal with
this, \peace\/ combines our own version of the $d^2$ alignment-free
sequence distance function \cite{Hide94} and the concept of a {\it
  minimum spanning tree} \cite{Prim57} to quickly and accurately find
clusters of ESTs expressed from the same gene without reference to a
sequenced genome.  Compared against clustering tool in the literature
\cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
  Pertea2003,Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be
of equivalent quality to the \wcd\/ and \tgicl tools
\cite{Hazelhurst08a,Pertea2003}, and more sensitive and robust than
other tools.  From the user perspective, o tool in the literature can
match \peace for ease of instillation, use, or the post-clustering
analysis tools \peace provides.

@@

While \peace\/ clearly matches \wcd\/ on long-read sequence, comparisons
against \tgicl\/ \cite{Pertea2003}  are more complicated.  In short, we find that when
parameters are picked appropriately, \tgicl\/ is more sensitive at lower
error rate -- but \peace\/ is more robust to error, in all cases
has an improved Type 2 error rates and is better able to
distinguish duplicated genes.  (See the Supplementary Materials for a
more detailed analysis.)

In applying the tools to real data Sanger data, we used the Human
Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
the A076941 {\it Arabidopsis thaliana} data set used to test wcd\/
\cite{Hazelhurst08a,Hazelhurst08b} (see Table~\ref{SangerReal}).  We
notice essentially identical results for \peace\/ and wcd\/ in
quality, both significantly better than \capthree\/ in Sensitivity and
Type 1 error rate, while slightly worse in the Jaccard Index and Type
2 error.  In runtime we see some inconsistency, with \peace\/ showing
a 60\% runtime improvement over \wcd\/ in the first data set, but
requiring 20\% more time than \wcd\/ in the Arabidopsis data set.  (We
cannot make a reasonable runtime comparison to \capthree, as the
runtime of that tool reflects both clustering and assembly.)  In
Table~S1 of the Supplementary Materials we present runtimes of several
more sets, observing that while \peace\/ appears to be significantly
faster on the smaller sets, \wcd\/ does overtake it for larger sets.

@@

We tested the three tools on short-read data using the \metasim\/ tool
of {\it Richter et al.} \cite{Richter2008}.  Encoded into \metasim\/
are sequence generation and error models corresponding to several
technologies, including the 454 short-read technology (producing reads
ranging in size from 200bp to 370bp with a mean of 250bp and a
standard deviation of 17bp), and the Illumina short-read technology
(producing reads of exactly 62bp).  In the 454 experiments \peace\/
achieved a sensitivity of 0.871, an 89.7\% improvement in sensitivity
over \wcd\/ and a 193\% improvement in sensitivity over \tgicl, with
comparable values for the Jaccard Index.  \wcd\/ was unable to handle
the Illumina data, while \peace\/ sensitivity, at a value of 0.384,
was nearly twice as good as Illumina.  (For more details, see Table~S3
in the Supplementary Materials.)  In short, \peace\/ performs quite
well for 454 reads and provides some useful information for Illumina
reads, with considerably better results than \wcd\/ or \tgicl\/.
